A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2020-04-23 paper/paper.Rnw
fix refs
Rename files to play nicely with auto-submission system

% Allow strikethrough text
\usepackage[normalem]{ulem}

\journal{Forensic Science International}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

%%% Commenting %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[table]{xcolor}
\newcommand{\mt}[1]{{\color{magenta} #1}}
\newcommand{\svp}[1]{{\color{teal} #1}}
\newcommand{\fix}[1]{{\color{orange} #1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<setup, include = F>>=
knitr::opts_chunk$set(echo = F, dev = "pdf", dpi = 1000, message = F, warning = F, fig.align = "center", fig.path = "figure/")

library(tidyverse)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(keras)

@@

\begin{document}

\begin{frontmatter}

\title{A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}}
\tnotetext[t1]{This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement \#70NANB15H176 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, University of California Irvine, and University of Virginia.}

%% Group authors per affiliation:
\author{Miranda Tilton and Susan Vanderplas}
\address{Center for Statistical Applications in Forensic Evidence, Iowa State University. 195 Durham Center, 613 Morill Rd., Ames, IA 50011}

% or include affiliations in footnotes:
% \author[1]{Miranda Tilton and Susan Vanderplas\corref{cor1}\fnref{isu}}
% \cortext[cor1]{Corresponding author}
% \ead{srvander@iastate.edu}
%
% \address[1]{195 Durham Center, 613 Morill Rd, Ames, IA 50011}
% \fntext[isu]{Center for Statistical Applications in Forensic Evidence, Iowa State University}

\begin{abstract}
Currently, to assess similarity in forensic shoeprint analysis, examiners must manually identify and label features in images of shoe prints or outsoles. In this paper, we discuss an algorithm for automatically labeling shoe outsole images using the same feature space examiners use: a set of features based on geometric shapes such as circles, stars, and quadrilaterals. The automatic labeling algorithm performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant improvement to the current manual classification of footwear tread patterns, and could be integrated into current matching databases to reduce the amount of time examiners spend on data entry.
\end{abstract}

\begin{keyword}
footwear, class characteristics, computer vision, neural networks
\end{keyword}

\end{frontmatter}

@@

\section{Introduction}
Most forensic examinations of footwear pattern evidence are limited to class characteristic based matches; that is, matches based on the positioning of geometric features within the tread pattern. Some of these features may include brand logos and size markings, while others are more generic geometric shapes, such as circles, triangles, stars, and chevrons. Class characteristics can be used for elimination, but are not sufficient for identification, as the characteristics are shared by all shoes with the same size, make, and model \citep{grossVariabilitySignificanceClass2013,bodziakfootwear2000}. Several attempts have been made to automatically identify features in both prints and shoe tread images, with low-level image analysis methods including fractal decomposition \citep{alexanderAutomaticClassificationRecognition1999}, scale-invariant feature recognition \citep{pavlouAutomaticExtractionClassification2006}, Fourier-Mellin transformation \citep{guehamAutomaticRecognitionShoeprints2008}, and other classical image analysis methods. Some of these low-level methods perform relatively well in good conditions \citep{luostarinenMeasuringAccuracyAutomatic2014a} but have degraded performance under conditions which are suboptimal, including those commonly found at crime scenes.

In the wider field of computer vision, attention has turned to more robust methods for image recognition, such as convolutional neural networks (CNNs), which are capable of achieving near-human accuracy under even degraded image conditions \citep{krizhevskyImageNetClassificationDeep2012}. These networks are designed to mimic the process of human vision, and typically involve application of sets of filters, some of which mimic the filters used in the low-level techniques used in early automatic classification attempts. CNNs have been applied in footwear forensics, making use of more general neural networks optimized to detect objects found in natural scenes, such as trees, animals, buildings, and cars; the pre-trained network's features are then used to compare shoes or prints to determine how well they match \citep{kongCrossDomainImageMatching2019a,kongCrossDomainForensicShoeprint2017,zhangAdaptingConvolutionalNeural2017}. Neural network methods appear to be more successful than traditional image analysis techniques, but the features used for matching are not generally informative for humans, who typically classify shoes by patterns and spatial relationships \citep{davisIntelligenceApproachFootwear1981,grossVariabilitySignificanceClass2013}.

In this paper, we discuss an alternate approach that uses an additional model layer to transfer the feature vector output into a vector of probabilities representing the detection of geometric elements in shoe tread images. Working within the feature space used by forensic examiners allows us to augment human-identified features with model output, assessing similarity of different shoe images on a feature set that is explicitly relevant to the domain. In addition, much of the existing software to assess geometric similarity, such as SoleMate\footnote{\url{http://www.fosterfreeman.com/trace-evidence/357-sicar-6-solemate-2.html}}, requires examiners to identify and label features manually; automating classification within the currently used feature space could reduce or eliminate the need for such manual classification.

\section{Materials and Methods}
%\fix{We may want to mention some of the methods in the introduction for continuity sake...
Ultimately, most pattern evidence is stored in photographic form; as a result, we must use image analysis methods to create numerical features from the original visual record. There are a number of methods that may be employed to identify shapes and features in an image, such as Fourier-Mellin transforms, Hough transformations \citep{ballardGeneralizingHoughTransform1981} and other low-level feature extraction methods aimed at detecting specific shapes, such as edges, corners, blobs, or ridges \citep[Ch 15]{machineVision}. While these methods are useful in identifying these specific features at a low level, they only identify features on a very small scale; as a result, they cannot reliably identify large geometric shapes like those that may be found in a high resolution outsole image.

@@

@@

\begin{figure}
\centering
\includegraphics[width=.1\textwidth]{images/augmentation/bowtie(R)-1-ugg-sienna-enamel-blue_product_8726365_color_120557.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_bowtie(R)-1-ugg-sienna-enamel-blue_product_8726365_color_120557_0_8344.jpg}\hfill\hfill\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/quad-2-nike-kids-downshifter-7-infant-toddler-gunsmoke-sunset-pulse-atmosphere-grey_product_8800875_color_720973.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_quad-2-nike-kids-downshifter-7-infant-toddler-gunsmoke-sunset-pulse-atmosphere-grey_product_8800875_color_720973_0_7533.jpg}\hfill\hfill\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/quad-2-puma-suede-classic-x-mac-three-port-royale-port-royale_product_9123838_color_251426.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_quad-2-puma-suede-classic-x-mac-three-port-royale-port-royale_product_9123838_color_251426_0_5689.jpg}\hfill\hfill\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/text-1-bernie-mev-kids-catwalk-little-kid-big-kid-multi-camo_product_9084647_color_87089.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_text-1-bernie-mev-kids-catwalk-little-kid-big-kid-multi-camo_product_9084647_color_87089_0_1891.jpg}
\caption[Original and augmented images.]{Four sets of original (left) and augmented (right) labeled images.}\label{fig:augmented}
\end{figure}

@@

We have proposed several hypotheses to explain why the model's predictions are not entirely consistent with manual labels on the images. We can test these hypotheses by examining what the model ``sees" when making its predictions. Heatmaps, also known as Class Activation Maps (CAMs), are a visual diagnostic tool that highlights the areas of an image that are most significant to classification for a given class \citep{chollet_allaire_2018}. Using these graphics, we can identify which locations in an image contribute most to the model's predictions.

@@

\section{Conclusions}
% Restate goal and briefly summarize methods
The goal of this research was to develop a method to automatically identify geometric class characteristics of shoe outsoles. A set of geometric class characteristics was defined, based on feature sets currently used by examiners, to both broadly classify a large variety of shoes and to narrow down similarity into a manageable number and type of features for further use in a shoeprint analysis. Thousands of outsole images were obtained from online shoe retailers and labeled according to their geometric shapes. The pre-trained convolutional base of VGG16 was then used, with a new classifier trained on a portion of these labeled images, to output predicted probabilities for test images within the defined set of geometric features. After training, the model performs well on the data provided. In general, the model is able to identify many well-defined geometric shapes in the images.

% Discussion of weaknesses
Although model predictions are generally accurate, there are a number of systematic misclassifications. The model has substantial difficulty predicting images with low contrast because image features are obscured. Exploring methods of color correction, such as histogram equalization, may prove useful for eliminating the effect of contrast on predictions. Initial implementations of histogram equalization suggest that the overall accuracy improvement is modest; the model fit to images which have been equalized had approximately 92.5\% validation accuracy, an improvement of 1.5\% over the default model described in this paper.

@@

Although there are many areas for potential improvement in future iterations of this model, we have demonstrated the effectiveness of transfer learning to provide an initial label for shoe outsole patterns. As most of the misclassifications are systematic, it is clear that the model is able to learn features and apply rules consistently. We expect that future modifications to the labeling criteria and data processing pipeline will improve the model's accuracy, but the 91\% validation accuracy is promising considering that we do not have estimates of the consistency of examiners' estimates on ambiguous features. Taken together, these factors show that Convolutional Neural Networks are a useful tool for classifying a large number of outsoles within a common classification scheme and can ultimately help streamline the current process of sorting and comparing shoes by class characteristics.

>>>>

\usepackage[normalem]{ulem} % Allow strikethrough text
\usepackage{float}
\usepackage{booktabs}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}

\usepackage{natbib}
\bibliographystyle{plain}

%%%%%%%%%%%%%%%%%%%%%%%

%%% Commenting %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \usepackage[table,dvipsnames]{xcolor}
\newcommand{\mt}[1]{{\color{magenta} #1}}
\newcommand{\svp}[1]{{\color{teal} #1}}
\newcommand{\fix}[1]{{\color{orange} #1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<setup, include = F>>=
knitr::opts_chunk$set(echo = F, dev = "pdf", dpi = 1000, message = F, warning = F, fig.align = "center", fig.path = "figure/")
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage[table]{xcolor}', x, fixed = TRUE)})

library(magrittr)
library(tidyverse)
library(tidyr)
nest <- nest_legacy
unnest <- unnest_legacy
library(ggplot2)
library(keras)

@@

\title{A Convolutional Neural Network for Outsole Recognition\footnote{This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement \#70NANB15H176 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, University of California Irvine, and University of Virginia.}}

%% Group authors per affiliation:
\author{Miranda Tilton and Susan Vanderplas}
% \address{Center for Statistics and Applications in Forensic Evidence, Iowa State University. 195 Durham Center, 613 Morill Rd., Ames, IA 50011}

\begin{document}
\maketitle
\begin{abstract}
Quantitative assessment of class characteristic frequency is a difficult problem, in part because there is currently no way to collect and efficiently process the quantity of data that would be required for such assessments without prohibitive amounts of human labor. In this paper, we discuss a computational method which complements a passive, unmanned scanning device that is currently under development. The automatic class characteristic recognition algorithm described in this paper has been developed with photographs of outsole images, and identifies high-level geometric features such as circles, stars, and quadrilaterals. These features are designed to mimic high-level features used by examiners to facilitate automatic processing of features without resorting to black-box feature sets commonly used in neural networks. The algorithm performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant step toward enabling the collection and application of population-level class characteristic data to support quantitative assessments of the value of class-characteristic based similarity assessments.
\end{abstract}
%
% \begin{keyword}
% Footwear; class characteristics; \mt{random match probability}; computer vision; neural networks
% \end{keyword}

% \end{frontmatter}

@@

\section{Introduction}\label{sec:Introduction}

Collection and use of footwear evidence varies heavily across jurisdictions within the United States, and, though it is more common for such analyses to be used in Europe, the U.K., and Israel, many of the same fundamental issues remain. It is often difficult to recover footwear evidence with sufficient detail to identify individualizing characteristics, either because of degradation or because the medium did not record the individualizing characteristics in sufficient detail to examine. It is also common for prints to be only partially recorded or recovered, which further compounds the difficulty of performing these types of analysis. As a result, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. This tendency is also reflected in the SWGTREAD range of conclusions scale, which has two separate categories to indicate degrees of class characteristic based association out of a range of 7 conclusions\citep{SWGTREAD2013}. Currently, the guidelines also caution against any statistical assessment because accurate data and models to support statistical calculations do not exist. However, if it were possible to collect a sufficiently large quantity of footwear impression data from a local population, it would be helpful to be able to calculate the frequency of shoes in the population that have similar class characteristics to a given shoe or print in order to characterize the determinative value of that particular piece of evidence. Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns on the class characteristic level.

Statistically, the coincidental (or random) match probability is estimated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print. The primary obstacle to coincidental match probability calculation, then, is one of data collection and annotation. The (limited) number of population class-characteristic studies \citep{parent_sandy_significance_nodate,benedict_geographical_2014} demonstrate the effort required to collect, process, and analyze even a moderate amount of data manually. However, technological innovations over the past decade have automated previously manual tasks such as license plate reading, tracking individuals on video feeds, and other complex tasks. In these comparable situations, two separate but related technological innovations were required: inexpensive and automatic data collection hardware (e.g. via CCTV or video cameras), and automatic, accurate data processing software. We are in the process of developing a low-profile scanner which, when embedded in a pedestrian area, passively collects images of shoe treads (a prototype is shown in \autoref{fig:prototype-scanner}). In parallel, however, it is necessary to develop software which can process the data from a similarly oriented scanner to generate features which are used to make class characteristic comparisons. Through the combination of hardware and software, we hope to enable the collection of population-level footwear data which will facilitate quantitative assessment of class characteristic frequency to support examiner evaluation.

\begin{figure}
\centering
\setlength\tabcolsep{1mm}
\begin{tabular}{cc}
\raisebox{-.5\height}{\includegraphics[width=.45\textwidth]{images/prototype/20200402_151639.jpg}} &
\raisebox{-.5\height}{\includegraphics[width=.45\textwidth]{images/prototype/GOPR1231-blur-enhance_square.jpg}} \vspace{1mm}\\

\end{tabular}
\caption[Images of prototype scanner.]{Images of the prototype scanner for collecting a large number of outsole images.}\label{fig:prototype-scanner}
\end{figure}

\subsection{Image Analysis}

% From original intro:
Several attempts have been made to automatically identify features in both prints and shoe tread images, with low-level image analysis methods including fractal decomposition \citep{alexanderAutomaticClassificationRecognition1999}, scale-invariant feature recognition \citep{pavlouAutomaticExtractionClassification2006}, Fourier-Mellin transformation \citep{guehamAutomaticRecognitionShoeprints2008}, and other classical image analysis methods. Some of these low-level methods perform relatively well in good conditions \citep{luostarinenMeasuringAccuracyAutomatic2014a} but have degraded performance under conditions which are suboptimal, including those commonly found at crime scenes.

% From original intro:
In the wider field of computer vision, attention has turned to more robust methods for image recognition, such as convolutional neural networks (CNNs), which are capable of achieving near-human accuracy under even degraded image conditions \citep{krizhevskyImageNetClassificationDeep2012}. These networks are designed to mimic the process of human vision, and typically involve application of sets of filters, some of which mimic the filters used in the low-level techniques used in early automatic classification attempts. CNNs have been applied in footwear forensics, making use of more general neural networks optimized to detect objects found in natural scenes, such as trees, animals, buildings, and cars; the pre-trained network's features are then used to compare shoes or prints to determine how well they match \citep{kongCrossDomainImageMatching2019a,kongCrossDomainForensicShoeprint2017,zhangAdaptingConvolutionalNeural2017}. Neural network methods appear to be more successful than traditional image analysis techniques, but the features used for matching are not generally informative for humans, who typically classify shoes by patterns and spatial relationships \citep{davisIntelligenceApproachFootwear1981,grossVariabilitySignificanceClass2013}.

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner suggest that cameras are more robust than impression-based systems (e.g.  EverOS) for use in outdoor settings. As a result, in this paper describing an initial automated software system for class characteristic recognition, we use outsole images (such as those found on online retail sites) rather than impression-based training data. While we recognize that images from the scanner are likely to show some degradation compared to images found from online retailers, the fundamental features of the underlying machine learning models are similar. In this paper, we present a proof-of-concept model designed to identify class characteristics probabilistically, using labels similar to those used by examiners. We discuss the strengths and limitations of both the classification scheme and the neural-network based model, with important implications for future attempts to develop automatic high-level class characteristic recognition systems.

\section{Materials and Methods}

Images collected from a footwear scanner (as described in the previous section) would be stored in photographic form; as a result, we must use image analysis methods to create numerical features from the original visual record. There are a number of methods that may be employed to identify shapes and features in an image, such as Fourier-Mellin transforms, Hough transformations \citep{ballardGeneralizingHoughTransform1981} and other low-level feature extraction methods aimed at detecting specific shapes, such as edges, corners, blobs, or ridges \citep[Ch 15]{machineVision}. While these methods are useful in identifying these specific features at a low level, they only identify features on a very small scale; as a result, they cannot reliably identify large geometric shapes like those that may be found in a high resolution outsole image.

@@

A reasonable critique of this approach is that images collected from online retailers are of much higher quality than would be expected from the sidewalk scanner proposed in \autoref{sec:Introduction}, and thus the CNN described in this paper is not immediately ready to process real degraded images. This consideration is justified, as machine learning methods tend to perform poorly when applied to test data that do not resemble the training data. However, the use of CNNs to predict class characteristics using the same categories used by human examiners is novel and introduces some interesting complications. As a full-featured automatic scanner is still under development, obtaining representative degraded images is not possible (because the scanner's optics are still in flux). The goal of this work is to evaluate the proposed classification scheme and determine whether CNNs can sufficiently distinguish class characteristic features found on outsoles. Addressing these questions with clean retailer images provides justification for additional algorithm development as well as continuing investment in optimization of the scanner design.

@@

\begin{figure}
\centering
\includegraphics[width=.1\textwidth]{images/augmentation/bowtie_orig.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_bowtie.jpg}\hfill\hfill\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/quad-2_orig.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_quad-2.jpg}\hfill\hfill\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/quad_orig.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_quad.jpg}\hfill\hfill\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/text_orig.jpg}\hfill
\includegraphics[width=.1\textwidth]{images/augmentation/aug_text.jpg}
\caption[Original and augmented images.]{Four sets of original (left) and augmented (right) labeled images.}\label{fig:augmented}
\end{figure}

@@

We have proposed several hypotheses to explain the inconsistencies between the manual labels and the model's predictions on the images. We can test these hypotheses by examining what the model ``sees" when making its predictions. Heatmaps, also known as Class Activation Maps (CAMs), are a visual diagnostic tool that highlights the areas of an image that are most significant to classification for a given class \citep{chollet_allaire_2018}. Using these graphics, we can identify which locations in an image contribute most to the model's predictions.

@@

\section{Conclusions}
% Restate goal and briefly summarize methods

In this paper, we have presented a method for automatic identification of geometric class characteristics in outsole tread patterns. A set of geometric class characteristics were defined based on feature sets currently used by examiners and used to classify a large variety of tread patterns. Labeling these images provided the opportunity to select a manageable number of features and define criteria for labeling these features on newly collected footwear images. Using the convolutional base of VGG16, a pre-trained neural network, we created a new classifier optimized to work with outsole images. This classifier returns the probability of each geometric feature for new outsole images.

% Discussion of weaknesses
Although model predictions are generally accurate, there are a number of systematic misclassifications and weaknesses which can be addressed in future iterations of the modeling process. The model has substantial difficulty predicting images with low contrast because image features are obscured. Exploring methods of color correction, such as histogram equalization, may prove useful for eliminating the effect of contrast on predictions. Initial implementations of histogram equalization suggest that the overall accuracy improvement is modest; the model fit to images which have been equalized had approximately 92.5\% validation accuracy, an improvement of 1.5\% over the default model described in this paper.

@@

Although there are many areas for potential improvement in future iterations of this model, we have demonstrated the effectiveness of using transfer learning to provide an initial label for shoe outsole patterns. As most of the misclassifications are systematic, it is clear that the model is able to learn features and apply rules consistently. We expect that the proposed modifications to the labeling criteria and data processing pipeline will improve the model's accuracy, but the 91\% validation accuracy is promising considering that we do not have estimates of the consistency of examiners' estimates on ambiguous features. Taken together, these factors show that Convolutional Neural Networks are a useful tool for classifying a large number of outsoles within a common classification scheme, although further research is needed into designing human comphrehensible classification schemes for this purpose. This work thus provides the necessary proof-of-concept that a well-trained model within an appropriate classification scheme, in conjunction with images collected from a footwear scanner, can be used to estimate the coincidental match probability of footwear tread patterns.
